Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-02-28 thesis.tex
reduce num headers in TOC
fix long list of names in citation
todolist at end for ease of finding
write most of intro

% Thesis-specific maths commands
\newcommand{\dmax}{d_\text{max}}
\newcommand{\psens}{p_\text{sens}}
\newcommand{\ntot}{n_\text{tot}}
\newcommand{\ndet}{n_\text{d}}
\newcommand{\nnodet}{n_\text{u}}
\newcommand{\pnodet}{p_\text{u}}
\newcommand{\Npop}{N_\text{pop}}
\newcommand{\Ncis}{N_\text{CIS}}
\newcommand{\ncis}{\vec{n_\text{CIS}}}
\newcommand{\na}{\vec{n_\text{a}}}
\newcommand{\pcis}{\vec{p_\text{CIS}}}
\newcommand{\sched}{\mathbb{T}}
\newcommand{\rate}{\mathfrak{R}}
\newcommand{\R}{\mathcal{R}_{0}}
\newcommand{\vlod}{v_\text{lod}}
\newcommand{\inform}{{_{\text{inform}}}}

@@

%% Bibliography
\usepackage[backend=biber,uniquename=false,sortcites=true,sorting=ynt,maxcitenames=2,style=apa]{biblatex}
\addbibresource{references.bib}
\addbibresource{refs-custom.bib}
% Display a linked DOI
\DeclareFieldFormat{doi}{%
  \mkbibacro{DOI}\addcolon\space
  \ifhyperref
    {\href{https://doi.org/#1}{\nolinkurl{#1}}}
    {\nolinkurl{#1}}%
}

@@

%% REMOVE THESE FOR FINAL VERSION WITHOUT TODOS
\usepackage{geometry}
\setlength{\marginparwidth}{3cm}
\geometry{margin=3.8cm}

@@

\chapter{Conclusion}
Household structure is not included in this model.
The household-based design of the survey is likely to induce clustering in results, increasing the variance of the measured prevalence.
The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.

\listoftodos

@@

\subfile{distributions}
\subfile{ATACCC-appendix-original-analysis}
\subfile{transmission-appendix-INLA}
\subfile{transmission-appendix-phase-type}

\todo[inline]{Delete below lines - only here to make compiling individual chapters work}
\cref{perf-test:eq:piu}
\cref{inc-prev:eq:forward-substitute}
\end{document}

>>>>

% Thesis-specific maths commands
\newcommand{\dmax}{d_\text{max}}
\newcommand{\psens}{p_\text{sens}}
\newcommand{\ntot}{n_\text{tot}}
\newcommand{\ndet}{n_\text{d}}
\newcommand{\nnodet}{n_\text{u}}
\newcommand{\pnodet}{p_\text{u}}
\newcommand{\Npop}{N_\text{pop}}
\newcommand{\Ncis}{N_\text{CIS}}
\newcommand{\ncis}{\vec{n_\text{CIS}}}
\newcommand{\na}{\vec{n_\text{a}}}
\newcommand{\pcis}{\vec{p_\text{CIS}}}
\newcommand{\sched}{\mathbb{T}}
\newcommand{\rate}{\mathfrak{R}}
\newcommand{\R}{\mathcal{R}_{0}}
\renewcommand{\Re}{\mathcal{R}_{e}(t)}
\newcommand{\Rc}{\mathcal{R}_{c}(t)}
\newcommand{\vlod}{v_\text{lod}}
\newcommand{\inform}{{_{\text{inform}}}}

@@

%% Bibliography
\usepackage[backend=biber,uniquename=init,sortcites=true,sorting=ynt,maxcitenames=2,style=apa,uniquelist=false]{biblatex}
\addbibresource{references.bib}
\addbibresource{refs-custom.bib}
% Display a linked DOI
\DeclareFieldFormat{doi}{%
  \mkbibacro{DOI}\addcolon\space
  \ifhyperref
    {\href{https://doi.org/#1}{\nolinkurl{#1}}}
    {\nolinkurl{#1}}%
}

@@

\setcounter{tocdepth}{1}

%% REMOVE THESE FOR FINAL VERSION WITHOUT TODOS
% \usepackage{geometry}
% \setlength{\marginparwidth}{3cm}
% \geometry{margin=3.8cm}

@@

\chapter{Conclusion} \label{conclusion}
Household structure is not included in this model.
The household-based design of the survey is likely to induce clustering in results, increasing the variance of the measured prevalence.
The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.

@@

\subfile{distributions}
\subfile{MCMC-appendix}
\subfile{ATACCC-appendix-original-analysis}
\subfile{transmission-appendix-INLA}
\subfile{transmission-appendix-phase-type}

\todo[inline]{Delete below lines - only here to make compiling individual chapters work}
\cref{perf-test:eq:piu}
\cref{inc-prev:eq:forward-substitute}
\listoftodos

\end{document}

### 2024-02-28 transmission.tex
write most of intro

I compute the posterior distribution using MCMC.
Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (previously discussed in \cref{E-inc-prev:sec:MCMC}).
The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
See \cref{SEIR:MCMC-algorithm} for full details.
\begin{algorithm}
 set $\vec{X_0}$ to an initial value of the parameter vector \;
 $\vec{\mu_0} = \vec{X_0}$ \;
 $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
 $\lambda_0 = 1$ \;
 \For{$i = 1, \dots, M$}{
  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
  update the proportion of proposals accepted so far, $\alpha$ \;
  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
    $\lambda_i = \lambda_{i-1}$ \;
    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
   }{
    $\gamma_i = (i - 200)^{-0.6}$ \;
    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
  }
 }
 \caption{Algorithm for adaptive random-walk Metropolis-Hastings. $\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$. $\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$. $\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%). $\gamma_i$ is the learning rate, which determines how much adaptation occurs. $\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}. Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
 \label{SEIR:MCMC-algorithm}
\end{algorithm}

@@

All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-inc-prev:sec:MCMC}).
The lowest Rhat was 1.01 for $\sigma_\epsilon$ in London.
All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.

>>>>

I compute the posterior distribution using MCMC.
Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (discussed in \cref{E-MCMC}).
The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
See \cref{SEIR:MCMC-algorithm} for full details.
\begin{algorithm}
 set $\vec{X_0}$ to an initial value of the parameter vector \;
 $\vec{\mu_0} = \vec{X_0}$ \;
 $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
 $\lambda_0 = 1$ \;
 \For{$i = 1, \dots, M$}{
  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
  update the proportion of proposals accepted so far, $\alpha$ \;
  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
    $\lambda_i = \lambda_{i-1}$ \;
    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
   }{
    $\gamma_i = (i - 200)^{-0.6}$ \;
    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
  }
 }
 \caption{Algorithm for adaptive random-walk Metropolis-Hastings. $\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$. $\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$. $\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%). $\gamma_i$ is the learning rate, which determines how much adaptation occurs. $\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}. Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
 \label{SEIR:MCMC-algorithm}
\end{algorithm}

@@

All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-intro:sec:MCMC}).
The lowest Rhat was 1.01 for $\sigma_\epsilon$ in London.
All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.
